---
title: "From CASHomon model sets to feature importance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From CASHomon model sets to feature importance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8, fig.height = 5)
library(cashomon)
```

## A set of good models

`cashomon` searches a finite pool of algorithms and hyperparameter settings using
TruVaRImp, Algorithm 1 in [Ewald et al. (2026)](https://arxiv.org/pdf/2603.15321).
Its goal is to find several configurations with similar predictive performance.
Comparing their explanations shows how much feature reliance can vary among
those models.

The numerical search works without the optional ML packages:

```{r numerical-search}
domain <- data.frame(x = seq(0, 1, length.out = 12))
search <- truvarimp(domain, function(row) (row$x - 0.3)^2,
  epsilon_abs = 0.05, noise = 0, budget = 8)
search
```

`L`, `H` and `U` contain indices predicted inside, outside and unresolved with
respect to the implicit loss threshold. `M` contains potential minimizers.
These are surrogate classifications; `fit_cashomon()` verifies the fitted
models before returning the empirical set.

## Training, selection and explanation data

Reserve three disjoint sets of rows: training for fitting candidates, validation
for selecting near-optimal models, and a holdout for PFI. This prevents explanation
data from influencing membership.

The following sections require `mlr3`, `rpart`, `xplainfi` and `R6`. They run when
these optional packages are installed; otherwise their code remains visible.

```{r optional-dependencies, include=FALSE}
can_fit <- all(vapply(c("mlr3", "rpart"), requireNamespace, logical(1), quietly = TRUE))
can_explain <- can_fit && all(vapply(c("xplainfi", "R6"), requireNamespace,
  logical(1), quietly = TRUE))
can_plot <- can_explain && requireNamespace("ggplot2", quietly = TRUE)
```

```{r data, eval=can_fit}
task <- mlr3::as_task_regr(mtcars, target = "mpg")
set.seed(42)
rows <- sample(task$row_ids)
train <- rows[1:16]
validation <- rows[17:24]
test <- rows[25:32]
```

This small dataset demonstrates the API. Use larger, appropriately stratified
or grouped partitions for substantive analysis.

## Search and verify the model set

```{r fit-models, eval=can_fit}
pool <- cashomon_candidates(
  learners = list(Tree = mlr3::lrn("regr.rpart"),
                  Baseline = mlr3::lrn("regr.featureless")),
  grids = list(Tree = list(cp = c(0.001, 0.01, 0.05), maxdepth = c(2L, 4L)))
)
models <- fit_cashomon(task, pool, train, validation,
  epsilon_rel = 0.20, budget = 5, verify_all = TRUE)
models
models$members
```

The empirical threshold is the best evaluated validation loss multiplied by
`1 + epsilon_rel`, plus `epsilon_abs`. All returned fitted models satisfy it.
`verify_all = TRUE` evaluates the remaining pool after active search, making
membership exact for this small finite pool. These extra fits are counted in
`n_verification` and are outside the search budget.

For larger pools, `verify_all = FALSE` verifies unfitted members of `L` and uses
the already evaluated candidates. Unobserved configurations may then be missing
from the returned set, or may have a better loss than the empirical reference.
Neither mode guarantees population-risk performance.

## Calculate PFI for every retained model

```{r pfi, eval=can_explain}
pfi <- cashomon_pfi(models, test_rows = test, n_repeats = 30, seed = 42)
head(pfi$importance)
summarise_cashomon_pfi(pfi, by_class = TRUE)
```

`pfi$importance` has one row per fitted model and feature. `pfi$scores` retains
each permutation. The adapter preserves the actual selected predictor during
xplainfi computation, including stochastic fits.

## Compare feature reliance

```{r cloud, eval=can_plot}
plot_cashomon_pfi(pfi, "cloud")
```

Each point is a model's mean PFI. The bands show the full range across retained
models and diamonds mark the median. Use `permutation_intervals = TRUE` to add
within-model permutation percentiles; these are not confidence intervals.

```{r heatmap, eval=can_plot}
plot_cashomon_pfi(pfi, "heatmap")
```

The heatmap groups model profiles by class and orders them by validation loss.
Negative importance values remain visible.

```{r ranks, eval=can_plot}
plot_cashomon_pfi(pfi, "ranks")
```

Ranks show whether models agree on the most important features. Tied importances
receive average ranks.

```{r performance, eval=can_plot}
plot_cashomon_pfi(pfi, "performance")
```

The performance view compares PFI against validation loss, with the membership
threshold marked. Each returned plot is a ggplot object that can be customized
and saved with `ggplot2::ggsave()`.

Across-model ranges describe only the retained finite set. Permutation variation
and variation across models are distinct; permutations are averaged within each
model before summarizing across models. Marginal PFI can also be affected by
correlated features and unusual combinations produced by permutation.

The installed regression example adds linear models and random forests and
exports tables and PNG/PDF plots:

```{r installed-example, eval=FALSE}
source(system.file("examples", "workflow.R", package = "cashomon"))
```
